---
title: "Introduction to SAS, Working with a mix of categorical and continuous variables"
author: "Steve Simon"
date: "Created: 2019-07-19"
output: powerpoint_presentation
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../results", output_format = "all") })  
---

```{r common-code, child="common-functions.Rmd"}
```

```{r}
file_prefix <- "5507-05-simon-working-with-mix-of-variables"
sas_file <- paste0(file_prefix, ".sas")
tx <- readLines(sas_file)
```

```{r show-general-structure}
tx %>%
  str_subset("^\\* ") %>%
  paste(collapse="\n\n") -> general_structure
tx %<>% str_trim(side="right") %>% paste0(collapse="\n") %>% str_split("\n\n\n") %>% unlist
output_page <- 0
```

### Overview

+ Review
  + proc format, proc freq, proc means
+ proc corr
+ proc sgplot
  + scatterplot
  + boxplot
+ by statement

<div class="notes">

author: Steve Simon
date: Created 2021-05-30
purpose: to produce slides for module05 videos
license: public domain

Here is an overview of what I want to cover in module05. 

</div>

`r add_chunks(0:5, tx)`

### Break #1
+ What have you learned
  + Reviewing descriptive statistics
+ What's next
  + Correlations and scatterplots

`r add_chunks(6:8, tx)`

### Break #2
+ What have you learned
  + Correlations and scatterplots
+ What's next
  + Boxplots
  
`r add_chunks(9:9, tx)`
  
### Break #3
+ What have you learned
  + Boxplots
+ What's next
  + Means by group

`r add_chunks(10:10, tx)`

### Break #4
+ What have you learned
  + Means by group
+ What's next
  + Investigating an odd association

`r add_chunks(11:11, tx)`

### Summary
+ Reviewing descriptive statistcs
+ Correlations and scatterplots
+ Boxplots
+ Means by group
+ Investigating an odd association


```{r quality-check}
fn <- "../results/qc05_chunks.txt"
qc <- "QC check of chunks\n"
for (i in 0:11) {
  qc %<>% append("\n<!--Chunk " %0% i %0% "-->\n")
  qc %<>% append(add_chunks(i:i, tx))
}
writeLines(qc, con=fn)
fn <- "../results/qc05_tx.txt"
qc <- "QC check of tx\n"
for (i in 1:length(tx)) {
  qc %<>% append("\n<!--Tx " %0% i %0% "-->\n")
  qc %<>% append(tx[i])
}
writeLines(qc, con=fn)
```