---
title: "Running a basic SAS program"
author: "Steve Simon"
date: "Created 2021-05-30"
output: powerpoint_presentation
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, comment="")
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(stringr)))
```

### Basic program, part 1

```{r}
tx <- readLines("m01-5507-simon-basic-program.sas")
tx %>% 
  str_subset(fixed("*")) %>%
  paste0(collapse="\n") -> tx1  
```

```{}
`r tx1`
```

<div class="notes">

author: Steve Simon and [put your name here]
date: created 2021-05-30
purpose: to produce slides for module01 videos
license: public domain

Here is the first piece of the basic program I want you to run. It has a brief documentation header that lists the author, date, purpose of the program, and license. I always place my code in the public domain.

</div>

### Basic program, part 2

```{r}
tx %>% str_which("^data") -> i1
tx %>% str_which("^;") -> i2
tx[i1:i2] %>%
  paste0(collapse="\n") -> tx2  
  
```

```{}
`r tx2`
```

<div class="notes">



</div>

### Break #1

+ What have you learned.

+ What's coming next



### Summary

+ What have you learned
