---
title: "Introduction to SAS, Working with categorical variables"
author: "Steve Simon"
date: "Created: 2021-05-30"
output: powerpoint_presentation
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../results", output_format = "all") })  
---


```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, comment="")
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(stringr)))
```

### Overview

+ proc format
+ recoding
+ proc freq
+ barcharts

<div class="notes">

author: Steve Simon
date: created 2021-05-30
purpose: to produce slides for module04 videos
license: public domain

Here is an overview of what I want to cover in this module.

We're going to look at a different data set, one with mostly categorical variables. I'll introduce proc format, which allows you to attach labels to categorical data, talk about recoding, and show some tables using proc freq. I'll also show you a simple bar chart.

</div>

```{r}
tx <- readLines("m04-5507-simon-categorical-variables.sas")

zpad <- function(i) {ifelse(i <10, paste0("0", i), paste0(i))}
```

```{r}
px <- function(tx, s1, s2, d1=0, d2=-2) {
  tx %>% str_which(s1) -> i1
  tx %>% str_which(s2) -> j1
  if (length(i1)==0) return("No beginning found")
  if (length(j1)==0) return("No ending found")
  if (length(i1) > 1) return("Multiple beginnings found")
  if (length(j1) > 1) return("Multiple endings found")
  i1 <- i1 + d1
  j1 <- j1 + d2
  tx %>%
    extract(i1:j1) %>%
    return
}
```

```{r}
co <- function(tx, i, d1=2, d2=-2) {
  s1 <- paste0("Part", zpad(i))
  s2 <- paste0("Notes", zpad(i))
  px(tx, s1, s2, d1, d2) %>%
    append("```{}", after=0) %>%
	append("```") %>%
    paste0(collapse="\n") %>%
	return
}
```

```{r}
ap <- function(tx) {
  '\n<div class="notes">\n' %>%
    append(tx) %>%
	append('\n</div>\n') %>%
	paste0(collapse="\n")
}

no <- function(tx, i, d1=0, d2=-2, part=0) {
  s1 <- paste0("Notes", zpad(i))
  s2 <- paste0("Part", zpad(i+1))
  px(tx, s1, s2, d1, d2) %>%
    str_replace(fixed("* "), "") %>%
  	str_replace(";$", "") -> notes_text
  if (part==0) 
    return(ap(notes_text))
  blank_lines <- c(1, str_which(notes_text, "^\\s*$"), length(notes_text))
  if (part > length(blank_lines)-1)
    return(paste0(c("Error", blank_lines), collapse=","))
  ra <- blank_lines[part]:blank_lines[part+1] 
  return(ap(notes_text[ra]))
}
```

```{r}
ti <- function(tx, i) {
  s1 <- paste0("Part", zpad(i))
  px(tx, s1, s1, 0, 0) %>%
  	str_replace(fixed("* Part"), "") %>%
	  str_replace("^....", "") %>%
    return   
}
```

```{r}

`%0%` <- function(a, b) {paste0(a, b)}

ou <- function(j) {
  "![SAS Output]" %0%
    "(../images/m04-5507" %0%
	"-simon-categorical_" %0%
	zpad(j) %0%
	".png)"
}
```

### SAS code: `r ti(tx, 0)` <!--documentation header-->
`r co(tx, 0)`
`r no(tx, 0)`


### SAS code: `r ti(tx, 1)` <!--where to find and store things-->
`r co(tx, 1)`
`r no(tx, 1)`


### SAS code: `r ti(tx, 2)` <!--import-->
`r co(tx, 2)`
`r no(tx, 2)`

### SAS code: `r ti(tx, 3)` <!--print-->
`r co(tx, 3)`
`r no(tx, 3, part=1)`

### SAS output: `r ti(tx, 3)`
`r ou(1)`
`r no(tx, 3, part=2)`

### SAS code: `r ti(tx, 4)` <!--freq-->
`r co(tx, 4)`
`r no(tx, 4, part=1)`

### SAS output: `r ti(tx, 4)` (1/2)
`r ou(2)`
`r no(tx, 4, part=2)`

### SAS output: `r ti(tx, 4)` (2/2)
`r ou(3)`
`r no(tx, 4, part=3)`

### SAS code: `r ti(tx, 5)`  <!--convert-->
`r co(tx, 5)`
`r no(tx, 5, part=1)`

### SAS output: `r ti(tx, 5)` (1/2)
`r ou(4)`
`r no(tx, 5, part=2)`

### SAS code: `r ti(tx, 6)` <!--format-->
`r co(tx, 6)`
`r no(tx, 6, part=1)`

### SAS output: `r ti(tx, 6)` 
`r ou(5)`
`r no(tx, 6, part=2)`

### SAS code: `r ti(tx, 7)` <!--bar chart-->
`r co(tx, 7)`
`r no(tx, 7, part=1)`

### SAS output: `r ti(tx, 7)` 
`r ou(6)`
`r no(tx, 7, part=2)`

### SAS code: `r ti(tx, 8)` <!--bar percentages-->
`r co(tx, 8)`
`r no(tx, 8)`

### SAS output: `r ti(tx, 8)` 
`r ou(7)`
`r no(tx, 8)`

### SAS code: `r ti(tx, 9)` <!--crosstabs-->
`r co(tx, 9)`
`r no(tx, 9, part=1)`

### SAS output: `r ti(tx, 9)` 
`r ou(8)`
`r no(tx, 9, part=2)`

### SAS code: `r ti(tx, 10)` <!-- -->
`r co(tx, 10)`
`r no(tx, 10)`

### SAS code: `r ti(tx, 11)` <!-- -->
`r co(tx, 11)`
`r no(tx, 11, part=1)`

### SAS output: `r ti(tx, 11)` 
`r ou(9)`
`r no(tx, 11, part=2)`

### SAS code: `r ti(tx, 12)` <!-- -->
`r co(tx, 12)`
`r no(tx, 12)`

### SAS code: `r ti(tx, 13)` <!-- -->
`r co(tx, 13)`
`r no(tx, 13)`

### SAS code: `r ti(tx, 14)` <!-- -->
`r co(tx, 14)`
`r no(tx, 14, part=1)`

### SAS output: `r ti(tx, 14)` 
`r ou(10)`
`r no(tx, 14, part=2)`

### SAS output: `r ti(tx, 14)` 
`r ou(11)`
`r no(tx, 14, part=3)`

### SAS code: `r ti(tx, 15)` 
`r co(tx, 15)`
`r no(tx, 15, part=1)`

### SAS output: `r ti(tx, 15)` 
`r ou(12)`
`r no(tx, 15, part=2)`

### SAS output:
`r ou(13)`
`r no(tx, 15, part=2)`

### SAS output:
`r ou(14)`
`r no(tx, 15, part=2)`

### SAS output:
`r ou(15)`
`r no(tx, 15, part=2)`

### Bar chart

Here are the descriptive statistics for age. Notice the number of missing values.

Here is what the output from proc freq looks like. Just two rows.

Here is what the output from proc freq looks like. Just two rows.

Here is what the output from proc freq looks like. Among the males, almost 5/6 died. Among the females only 1/3 died.

Here is the quality check. Notice that adult starts at 21. Should adult start at 18 instead?

Are the ranges for pre-teen and teenager reasonable?

How about the ranges for toddler?

This shows the age categories starting at the youngest: toddler and pre-teen...

followed y teenager and adult...

with missing bring up the rear. This order was the order of the number codes. So if you want to display your results in a non-alphabetical order, use number codes.

Here is the quality check. PClass=1st codes to first_class=Yes. PClass=2nd or 3rd codes to first_class=No.
