---
title: "Introduction to SAS, Working with continuous variables"
author: "Steve Simon"
date: "Created 2021-05-30"
output: 
  powerpoint_presentation: 
    reference_doc: ../results/wide-screen-template.pptx
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, 
  output_dir = "../results", output_format = "all") })  
---


```{r, echo=FALSE}
# Specify the SAS file you are parsing here.
fn <- "5507-02-simon-continuous-variables" 
```


```{r common-code, child="test-code.Rmd"}
```

### QC check

`r chunk_type`

### QC check

`r chunk_number`

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][1:5]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][6:10]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][11:15]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][16:20]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][21:23]
```


```{r echo=FALSE}
i <- 0
```

### Chunk `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

