---
title: "Introduction to SAS, Working with continuous variables"
author: "Steve Simon"
date: "Created 2021-05-30"
output: 
  powerpoint_presentation: 
    reference_doc: ../results/wide-screen-template.pptx
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, 
  output_dir = "../results", output_format = "all") })  
---

```{r setup, echo=FALSE}
suppressMessages(
  suppressWarnings(
    library(tidyverse)))
suppressMessages(
  suppressWarnings(
    library(glue)))
zpad <- function(i) {ifelse(i<10, paste0("0", i), paste0(i))}
```

```{r, echo=FALSE}
fn <- "5507-02-simon-continuous-variables" 
fn %>%
  paste0(".sas") %>%
  readLines %>%
  str_remove_all("^\\W*$") %>%
  paste0(collapse="\n") %>%
  str_split("\n\n\n") %>%
  unlist -> sas_text
```

```{r, echo=FALSE}
display_code <- function(slide_title, slide_text) {
  paste0(
    "### SAS Code: ", 
    slide_title,
    "\n\n```{}\n", 
    slide_text,
    "\n```\n\n")
}

display_notes <- function(slide_text) {
  paste0(
    '<div class="notes">', 
    "\n\n", 
    slide_text,
    "\n\n</div>\n\n")
}

display_output <- function(slide_title, slide_text) {
  paste0(
    "### SAS output: ", 
    slide_title,
    "\n\n", 
    slide_text,
    "\n\n")
}
```

```{r echo=FALSE}
backticks <- "```"

code_slide <- "

### SAS Code: {slide_title}

{backticks}{{}}

{sas_text[i]}

{backticks}

"
```

```{r echo=FALSE}
code_title <- "<!---{sas_text[i]}--->"
```

```{r echo=FALSE}
slide_notes <- '

<div class="notes">

{sas_text[i]}

</div>

'
```

```{r echo=FALSE}
output_slide <- '

### SAS output: {slide_title}

![Figure {output_page}. {sas_text[i]}](../images/{fn}_{zpad(output_page)}.png) 

'
```


```{r, echo=FALSE}
slide_title <- "### SAS code is not in the proper order"
chunk_type <- rep(99, length(sas_text))
output_page <- 0
for (i in 1:length(sas_text)) {
  chunk_type[i] <- 
    1*str_detect(sas_text[i], "^\\* Part") + 
    2*str_detect(sas_text[i], "^\\* Speaker notes") + 
    4*str_detect(sas_text[i], "^\\* Output from ") 
  
  if (chunk_type[i] > 0) {
    sas_text[i] <- str_remove(sas_text[i], "^\\* ")
    sas_text[i] <- str_remove(sas_text[i], ";$")
  }
  
  if (chunk_type[i] == 4) output_page <- output_page+1
  
  if (chunk_type[i]==1) {slide_title <- sas_text[i]}
  
  sas_text[i] <- case_when(
    chunk_type[i]==1 ~ as.character(glue(code_title)),
    chunk_type[i]==0 ~ as.character(glue(code_slide)),
    chunk_type[i]==2 ~ as.character(glue(slide_notes)),
    chunk_type[i]==4 ~ as.character(glue(output_slide)),
    TRUE ~ paste0("### Error: ", chunk_type[i])
  )
}

chunk_number <- cumsum(chunk_type==1)
```

=======
### QC check

`r chunk_type`

### QC check

`r chunk_number`

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][1:5]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][6:10]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][11:15]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][16:20]
```

### QC check

```{r echo=FALSE}
sas_text[chunk_type==1][21:23]
```


```{r echo=FALSE}
i <- 0
```

### Chunk `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

### Chunk  `r i <- i+1; i`

`r paste0(sas_text[chunk_number==i], collapse="\n\n")`

