---
title: "Introduction to SAS, Working with categorical variables"
author: "Steve Simon"
date: "Created: 2021-05-30"
output: powerpoint_presentation
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../results", output_format = "all") })  
---


```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, comment="")
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(stringr)))
```

### Overview

+ proc format
+ recoding
+ proc freq
+ barcharts

<div class="notes">

author: Steve Simon
date: created 2021-05-30
purpose: to produce slides for module04 videos
license: public domain

Here is an overview of what I want to cover in this module.

We're going to look at a different data set, one with mostly categorical variables. I'll introduce proc format, which allows you to attach labels to categorical data, talk about recoding, and show some tables using proc freq. I'll also show you a simple bar chart.

</div>

```{r}
tx <- readLines("m04-5507-simon-categorical-variables.sas")

zpad <- function(i) {ifelse(i <10, paste0("0", i), paste0(i))}
```

```{r}
px <- function(tx, s1, s2, d1=0, d2=-2) {
  tx %>% str_which(s1) -> i1
  tx %>% str_which(s2) -> j1
  if (length(i1)==0) return("No beginning found")
  if (length(j1)==0) return("No ending found")
  if (length(i1) > 1) return("Multiple beginnings found")
  if (length(j1) > 1) return("Multiple endings found")
  i1 <- i1 + d1
  j1 <- j1 + d2
  tx %>%
    extract(i1:j1) %>%
    paste0(collapse="\n") %>%
    return
}
```

```{r}
co <- function(tx, i) {
  s1 <- paste0("Part", zpad(i))
  s2 <- paste0("Notes", zpad(i))
  px(tx, s1, s2, 0, -2)
}
```

```{r}
no <- function(tx, i) {
  s1 <- paste0("Notes", zpad(i))
  s2 <- paste0("Part", zpad(i+1))
  px(tx, s1, s2, 0, -2) %>%
  	str_replace(fixed("* Notes"), "") %>%
  	str_replace(";$", "") %>%
	  str_replace("^[0-9][0-9]\\. ", "")
}
```

```{r}
ti <- function(tx, i) {
  s1 <- paste0("Part", zpad(i))
  px(tx, s1, s1, 0, 0) %>%
  	str_replace(fixed("* Part"), "") %>%
	  str_replace("^....", "") %>%
    return   
}
```

### SAS code: `r ti(tx, 0)`

```{}
`r co(tx, 0)`
```

<div class="notes">

`r no(tx, 0)`

</div>

